---
title: "Final Project - Robert Alvarez"
output: html_document
---

### Import useful libraries
```{r}
library(dplyr)
library(usmap)
library(ggplot2)
library(leaflet)
```

# 2.
## a. From trip.csv data","for the city San Francisco","please draw a network diagram to show the busiest bike riding routes on the map. Can you identify the top 2 busiest routes?
### Read trip.csv
```{r}
trips <- read.csv("trip.csv")
station <- read.csv("station.csv")
```

### Update trips zip_code values, so they match any of the 5 in weather by using the station ID in station
```{r}
#Get unique cities
cities <- unique(station$city)
#Make a data frame "map" from city to new zip code
city_to_zip_code <- data.frame(city=c("Mountain View","Palo Alto","Redwood City","San Francisco","San Jose"),
                               zip_code=c(94041, 94301, 94063, 94107, 95113))

#Update trips zip codes so they match one of the 5
for (city in cities) {
  #Get IDs of the current city
  IDs <- station[station$city == city, ] %>% subset(select=id)
  #Get new_zip_code
  new_zip_code <- city_to_zip_code[city_to_zip_code$city == city,][[2]]
  #Replace previous zip_code
  trips$zip_code[trips$start_station_id %in% IDs$id] <- new_zip_code
}
#trips
```

### Get only the trips that are from San Francisco and subset the start and end station
```{r}
SF_zip_code <- city_to_zip_code[city_to_zip_code$city == "San Francisco",][[2]]
SF_df <- trips[trips$zip_code == SF_zip_code,] %>% subset(select=c(start_station_id,end_station_id))
#SF_df
```

### Erase trips that start in San Francisco and end in another city
```{r}
SF_stations <- station[station$city == "San Francisco",]
SF_df <- SF_df[SF_df$end_station_id %in% SF_stations$id, ]
#SF_df
```

### Group start to end trips and count how many of each
```{r}
options(dplyr.summarise.inform = FALSE)
SF_routes <- SF_df %>% group_by(start_station_id,end_station_id) %>% summarise(n_trips = n()) %>% ungroup()
#SF_routes
```

### Get two most concurrent trips
```{r}
largest_idx <- which.max(SF_routes$n_trips)
largest <- SF_routes[largest_idx, ]
temp <- SF_routes[-largest_idx, ]
start <- station[station$id == largest$start_station_id, ]$name
end <- station[station$id == largest$end_station_id, ]$name
n_trips <- largest$n_trips
print(paste("Route with the most amount of trips is between", start, "and", end, "with", n_trips, "trips."))

largest_idx <- which.max(temp$n_trips)
largest <- temp[largest_idx, ]
start <- station[station$id == largest$start_station_id, ]$name
end <- station[station$id == largest$end_station_id, ]$name
n_trips <- largest$n_trips
print(paste("Second route with the most amount of trips is between", start, "and", end, "with", n_trips, "trips."))
```

### Change station IDs to its lat and long coordinates by using stations df
```{r}
SF_routes$start_lat <- 0
SF_routes$start_long <- 0
SF_routes$end_lat <- 0
SF_routes$end_long <- 0
for (ID in unique(station$id)) {
  SF_routes$start_lat[SF_routes$start_station_id == ID] <- station[station$id == ID,"lat"]
  SF_routes$start_long[SF_routes$start_station_id == ID] <- station[station$id == ID,"long"]
  SF_routes$end_lat[SF_routes$end_station_id == ID] <- station[station$id == ID,"lat"]
  SF_routes$end_long[SF_routes$end_station_id == ID] <- station[station$id == ID,"long"]
}
SF_routes <- SF_routes %>% subset(select=-c(start_station_id,end_station_id))
#SF_routes
```

### Make the plot
```{r}
#Make map with stations and labels
m <- leaflet() %>%
  setView(lng=(min(SF_stations$long)+max(SF_stations$long))/2, lat=(min(SF_stations$lat)+max(SF_stations$lat))/2, zoom=14) %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=SF_stations$long, lat=SF_stations$lat, label=SF_stations$name)

#Use created map to add lines representing routes
for (i in seq_len(nrow(SF_routes))){
  m <- addPolylines(m, lat = as.numeric(SF_routes[i, c(2, 4)]), lng = as.numeric(SF_routes[i, c(3, 5)]), weight=2)
}

#Plot map
m
```

## b. Repeat step (a) for the remaining cities.
### Make a function that plot a route map for each city in trips
```{r}
plot_routes <- function(trips, station, city_to_zip_code) {
  print_busiest_routes <- function(routes, station, city) {
    cat(paste0("For the city of ", city, ":"), "\n")

    largest_idx <- which.max(routes$n_trips)
    largest <- routes[largest_idx, ]
    temp <- routes[-largest_idx, ]
    start <- station[station$id == largest$start_station_id, ]$name
    end <- station[station$id == largest$end_station_id, ]$name
    n_trips <- largest$n_trips
    cat(paste("\tRoute with the most amount of trips is between", start, "and", end, "with", n_trips, "trips."), "\n")

    largest_idx <- which.max(temp$n_trips)
    largest <- temp[largest_idx, ]
    start <- station[station$id == largest$start_station_id, ]$name
    end <- station[station$id == largest$end_station_id, ]$name
    n_trips <- largest$n_trips
    cat(paste("\tSecond route with the most amount of trips is between", start, "and", end, "with", n_trips, "trips."), "\n")
  }
  #Get all 5 different zip codes
  cities <- city_to_zip_code$city

  #For each zip_code make a bike routes map
  for (city in cities) {
    zip_code <- city_to_zip_code[city_to_zip_code$city == city,][[2]]
    #Get trips that start from the city of interest
    city_df <- trips[trips$zip_code == zip_code,] %>% subset(select=c(start_station_id,end_station_id))
    #Erase trips that finish outside the city of interest
    city_stations <- station[station$city == city,]
    city_df <- city_df[city_df$end_station_id %in% city_stations$id, ]
    #Calculate amount of trips
    routes <- city_df %>% group_by(start_station_id,end_station_id) %>% summarise(n_trips = n()) %>% ungroup()
    #Print two busiest routes
    print_busiest_routes(routes, station, city)
    #Change IDs to lat and long
    routes$start_lat <- 0
    routes$start_long <- 0
    routes$end_lat <- 0
    routes$end_long <- 0
    for (ID in unique(station$id)) {
      routes$start_lat[routes$start_station_id == ID] <- station[station$id == ID,"lat"]
      routes$start_long[routes$start_station_id == ID] <- station[station$id == ID,"long"]
      routes$end_lat[routes$end_station_id == ID] <- station[station$id == ID,"lat"]
      routes$end_long[routes$end_station_id == ID] <- station[station$id == ID,"long"]
    }
    routes <- routes %>% subset(select=-c(start_station_id,end_station_id))
    #Make map with stations and labels
    m <- leaflet() %>%
      setView(lng=(min(city_stations$long)+max(city_stations$long))/2, lat=(min(city_stations$lat)+max(city_stations$lat))/2, zoom=14) %>%
      addTiles() %>%  # Add default OpenStreetMap map tiles
      addMarkers(lng=city_stations$long, lat=city_stations$lat, label=city_stations$name)

    #Use created map to add lines representing routes
    for (i in seq_len(nrow(routes))){
      m <- addPolylines(m, lat = as.numeric(routes[i, c(2, 4)]), lng = as.numeric(routes[i, c(3, 5)]), weight=2)
    }

    #Plot map
    print(m)
  }
}
```

## Make route plot for each city
```{r}
plot_routes(trips, station, city_to_zip_code)
```
